NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-03-12 main.tex
Update text

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} for MD simulation trajectories. The NMRlipids databank is constructed using the NMRlipids project protocol where all the content is openly accessible throughout the project~\cite{botan15}. Currently the NMRlipids databank contains 726 simulation trajectories with the total length of approximately 0.4\,ms. The distribution of available simulations containing a specific lipid, shown in Fig.~\ref{fig:overlay}B, roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. Abbreviations and full names of all lipids present in the databank are listed in Table~\ref{tab:abbreviations}. Single component lipid membranes and binary mixtures, depicted in Fig.~\ref{fig:overlay}D, are currently most abundant in the NMRlipids databank, yet mixtures with up to five lipid types are available. Force fields used in simulations cover all the essential parameters commonly used in lipid simulations, see Fig.~\ref{fig:overlay}C and table~\ref{tab:ForceFields}, including also united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

@@

The {\it application layer} is composed of repositories and tools that read information from the {\it databank layer} for further analyses. Because {\it application layer} does not interfere with the {\it databank layer}, it can be freely extended by anyone for wide range of purposes. Here we demonstrate this with two examples: NMRlipids databank graphical user interface (NMRlipids databank-GUI) tool at~\url{www.databank.nmrlipids.fi/} and a repository exemplifying novel analyses utilizing NMRlipids databank as discussed below (\url{www.github.com/NMRLipids/DataBankManuscript}). More detailed description of the NMRlipids databank structure is available in the supplementary information.

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{A) Schematic presentation of the overlay structure used in the NMRlipids databank. A more detailed structure of the databank layer is shown in Fig.~\ref{DatabankStructure} in the SI.
    %b) Distribution of the lengths of the trajectories, total number of trajectories and total length of the simulations in the NMRlipids databank.
    B) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or fewer simulations ('others') are listed on the right. 
    C) Distribution of force fields in the simulations in the NMRlipids databank. References for each force field are given in table~\ref{tab:ForceFields}.
    D) Currently available binary mixtures in the NMRlipids databank. 
    }
    \label{fig:overlay}
\end{figure}

\subsection{NMRlipids databank-GUI: graphical access to the MD simulation data}\label{section:access}
NMRlipids databank-GUI provides easy access to the content of the NMRlipids databank via graphical user interface (GUI) that is available at~\url{www.databank.nmrlipids.fi/}. The NMRlipids databank-GUI has advanced searching options for system content and properties. The search results are ranked based on their quality evaluation against experimental data which enables rapid selection of best available simulations for specific systems.  Systems can be also visualized and their properties can be compared with different simulations and experiments.

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, enables automatic analysis of virtually any property from all simulations in the NMRlipids databank. This enables wide range of novel analyses that take the advantage on unprecendented amount MD simulation data with programmatic access. This is realized in the NMRlipids databank-API by delivering functions that enable straightforward implementation of the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}A. Iteration through the README.yaml files in the {\it databank layer} enables the access to the raw data of each simulation within the databank via the links stored in these files. The affiliation of universal molecule and atom names to the simulation specific names are delivered for each simulation in README.yaml and mapping files, which enables automatic analyses over all systems independently on the specific naming convention used in a simulation. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

While analysis codes and results for basic membrane properties 
%area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness 
are included in the {\it Databank layer}, unlimited amount of further analyses can be implemented by anyone in separate repositories in the {\it application layer}. {\it Application layer} repositories that are organized using the {\it Databank layer} structure can be straightforwardly accessed programmatically and further analyzed by implementing the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}B by using the NMRlipids databank-API functions. Novel analyses demonstrating the power of the NMRlipids databank in selecting best simulation models, analyses of rare phenomena and extending MD simulations to new fields are implemented in an {\it application layer} repository located at~\url{www.github.com/NMRLipids/DataBankManuscript}.

@@

Divergent predictions for the lateral packing of POPC and POPE membranes from different force fields is illustrated in Fig.~\ref{fig:POPCPOPEapls}C. Because the membrane packing in terms of area per lipid correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F), simulations with the best quality for this part of POPE and POPC lipids are considered to give the most realistic results in Fig.~\ref{fig:POPCPOPEapls}C. To this end, the ranking based on C--H bond order parameter quality for the best performing POPC and POPE simulations in the NMRlipids databank are shown in Figs.~\ref{fig:quality}A and B, respectively. Direct comparisons with the experimental data are shown in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI} for the most relevant simulations. Based on these rankings, simulations with Slipids parameters are the most realistic for POPC lipids with a clear difference to others, and 2\textsuperscript{nd} most realistic for POPE with only marginally lower quality than GROMOS-CKP simulations. Simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evalution based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for the cases were membrane packing is relevant.

>>>>

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} for MD simulation trajectories. The NMRlipids databank is constructed using the NMRlipids project protocol where all the content is openly accessible throughout the project~\cite{botan15}. Currently the NMRlipids databank contains 726 simulation trajectories with the total length of approximately 0.4\,ms. The distribution of available simulations containing a specific lipid, shown in Fig.~\ref{fig:overlay}B, roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. Abbreviations and full names of all lipids present in the databank are listed in Table~\ref{tab:abbreviations}. Single component lipid membranes and binary mixtures, depicted in Fig.~\ref{fig:overlay}F, are currently most abundant in the NMRlipids databank, yet mixtures with up to five lipid types are available. Force fields used in simulations cover all the essential parameters commonly used in lipid simulations, see Fig.~\ref{fig:overlay}C and table~\ref{tab:ForceFields}, including also united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

@@

The {\it application layer} is composed of repositories and tools that read information from the {\it databank layer} for further analyses. Because {\it application layer} does not interfere with the {\it databank layer}, it can be freely extended by anyone for wide range of purposes. This is demonstrated here with two examples: NMRlipids databank graphical user interface (NMRlipids databank-GUI) tool at~\url{www.databank.nmrlipids.fi/} and a repository exemplifying novel analyses utilizing NMRlipids databank as discussed below (\url{www.github.com/NMRLipids/DataBankManuscript}). More detailed description of the NMRlipids databank structure is available in the supplementary information.

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{A) Schematic presentation of the overlay structure used in the NMRlipids databank. A more detailed structure of the databank layer is shown in Fig.~\ref{DatabankStructure} in the SI.
    %b) Distribution of the lengths of the trajectories, total number of trajectories and total length of the simulations in the NMRlipids databank.
    B) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or fewer simulations ('others') are listed on the right. 
    C) Distribution of force fields in the simulations in the NMRlipids databank. References for each force field are given in table~\ref{tab:ForceFields}.
    D) Flowchart for performing an analysis of properties through all MD simulations in the NMRlipids databank using the API.
    E) Flowchart for accessing results calculated from the NMRlipids databank and stored to the application layer.
    F) Currently available binary mixtures in the NMRlipids databank. 
    }
    \label{fig:overlay}
\end{figure}

\subsection{NMRlipids databank-GUI: graphical access to the MD simulation data}\label{section:access}
NMRlipids databank-GUI, available at~\url{www.databank.nmrlipids.fi/}, provides easy access to the NMRlipids databank content via graphical user interface (GUI). Simulations can be searched based on their composition, used force field, temperature, properties and quality. The search results are ranked based on simulation quality evaluated against experimental data. NMRlipids databank-GUI enables also visualization of simulations and comparison of properties between different simulations and experiments. The NMRlipids databank-GUI enables rapid surveys of available simulation data, selection of best available simulations for specific systems based on ranking lists, and comparisons of basic properties between different types of membranes. Notably, GUI enables these operations to be performed by scientists with wide ranges of backgrounds who do not necessarily have expertise in programming or other means to access to MD simulation data.

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, provides programmatic access to all simulation data in the NMRlipids databank. This enables wide range of novel data-driven applications from construction of machine learning models that predict membrane properties to automatic analyses of virtually any property from all simulations in the databank. 
%Wide range of novel analyses taking the advantage of programmatic access to unprecendented amount MD simulation data can be performed with the tools delivered by the NMRlipids databank-API. 
Practical implementation of such analyses is illustrated in the flowchart in Fig.~\ref{fig:quality}D. First, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can be then automatically analyzed with the help of README.yaml and mapping files that affiliate specific naming conventions in each simulation to the universal molecule and atom names. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

While analysis codes and results for basic membrane properties 
%area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness 
are included in the {\it Databank layer}, unlimited amount of further analyses can be implemented by anyone in separate repositories in the {\it application layer}. When {\it Application layer} repositories are organized by mimicking the {\it Databank layer} structure, they can be accessed programmatically and further analyzed using the tools in NMRlipids databank-API by implementing the flowchart demonstrated in Fig.~\ref{fig:quality}E. Novel analyses demonstrating the power of the NMRlipids databank in selecting best simulation models, analyses of rare phenomena and extending MD simulations to new fields are implemented in an {\it application layer} repository located at~\url{www.github.com/NMRLipids/DataBankManuscript}.

@@

Predictions for the lateral packing of POPC and POPE membranes in terms of area per lipid, one of the most important parameter used to characterize cellular membranes, diverge between different force fields as illustrated in Fig.~\ref{fig:quality}A. Because the membrane packing  correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F), simulations with the best quality for this part of POPE and POPC lipids are considered to give the most realistic results in Fig.~\ref{fig:POPCPOPEapls}C. To this end, the ranking based on C--H bond order parameter quality for the best performing POPC and POPE simulations in the NMRlipids databank are shown in Figs.~\ref{fig:quality}A and B, respectively. Direct comparisons with the experimental data are shown in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI} for the most relevant simulations. Based on these rankings, simulations with Slipids parameters are the most realistic for POPC lipids with a clear difference to others, and 2\textsuperscript{nd} most realistic for POPE with only marginally lower quality than GROMOS-CKP simulations. Simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evalution based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for the cases were membrane packing is relevant.
